Ubiquitous Digital Technologies and Spatial Structure; an update

### 2021-03-23 paper/2_final_submission/tranos_ioannides.Rmd
editing changes

Urban economics suggests that a key source of agglomeration externalities is proximity, which  facilitates interaction and knowledge spillovers [@rosenthal_evidence_2004].
Hence, face-to-face interactions and the implied knowledge spillovers within cities flourish because of lower transportation and interaction costs [@storper_buzz_2004].
However, ICT have the capacity to directly affect this process by further reducing transportation cost^[Or, more broader, spatial transmission costs to follow McCann’s terminology [@mccann_globalization_2008]] [@mccann_globalization_2008].
In essence, the internet and digital communications promote further learning and matching, which are at the heart of the micro-foundations of agglomeration economies identified by Duranton and Puga [@duranton_micro-foundations_2004].
Web-based applications such as Massive Open Online Courses decrease the need for co-location of actors in order to participate in formal learning activities.
Furthermore, online social media such as LinkedIn^[LinkedIn has been identified as "the largest professional matchmaker site in the world" [@van_dijck_you_2013], p. 207.
See also [@gee2019more].]
can enhance the probability of matching and indeed the quality of matches even within cities. Therefore, it is natural to ask whether the widespread adoption of ICT could offset the benefits of  immediate physical proximity and result in more dispersed spatial structures, or further reinforce agglomeration  externalities and lead to more concentrated spatial structures.

@@

<!-- <!-- \color{blue} --> -->
Kolko [@kolko_death_2000] uses internet diffusion data and identifies a clear complementary link between internet usage and city size.
Interestingly, he identifies higher internet domain densities in remote cities which indicates a substitution effect of the internet for longer-distance non-electronic communications. His results are consistent for different measures of internet diffusion (internet domain density and internet take-up).
Sinai and Waldfogel [@todd_sinai_geography_2004] approach the same question from the consumers’ point of view and find support for complementarity between internet and cities.
They study the link between market size and locally targeted online content and find more online local content in  larger markets.
However, they also obtain evidence for a substitution effect: holding local online content constant, market size has a negative effect on individual connectivity.
Forman et al. [@chris_forman_how_2005] examines whether commercial internet adoption is higher in cities than in rural areas. While the former would indicate a complementarity between internet adoption and cities, the latter would reflect a substitutability relation, according to which the internet is used as a means to offset costs and lack of opportunities related to peripheriality.
Their results indicate that despite internet adoption by firms with more than 100 employees being greater in smaller urban agglomerations, the adoption of more sophisticated internet-based applications is positively related with city size in $2000$.
Sohn, Kim, and Hewings [@jungyul_sohn_information_2003] compare how information technologies are related to urban spatial structure for Chicago and Seoul.
Although they find a clear complementary link for Chicago, this is not the case for Seoul, where information technologies contribute to a more dispersed spatial pattern.
Focusing on the municipalities in the province of Barcelona, Pons-Novell and Viladecans-Marsal [@pons-novell_cities_2006] find a complementary link between individual internet take-up and off-line commercial offerings, but their results cannot safely reject a substitution effect.
Craig, Hoang, and Kohlhase [@craig_does_2016] focus on internet take-up rates across US states during the period $2000-2011$.
Their analysis provides suggestive evidence of a complementary role that internet connectivity performs on urban living.
Ioannides et al. [@ioannides2008effect] use country-level data on city size distributions to examine the impact of fixed line telephony on urban structure $1980-2000$.
Using a panel dataset of spatial dispersion measures, they find robust evidence that an increase in the number of telephone lines per capita encourages the spatial dispersion of population in that they lead to a more concentrated distribution of city sizes.
In addition, by the end of their study period the internet has come into use, but their results with internet usage is more speculative but do show that it goes in the same direction.

<!-- <!-- \color{blue} --> -->
Focusing on rural areas, Partridge et al. [@mark_d_partridge_employment_2008] find no evidence that rural distance penalties in the US have substantially changed since $1970s$ indicating that technological changes including the internet and digital communications have not affected spatial structure. Also interesting are the findings of a more recent study by Kim and Orazem [@kim_broadband_2016] on the economic effects of broadband internet in rural areas. They identify a positive effect on new firm location decisions, with the effect being higher in rural areas with larger population and in rural areas adjacent to a metropolitan area. These results suggest a complementary relation between the internet and agglomeration economies.

<!-- <!-- \color{blue} --> -->
Moving beyond modelling the direct relationship between ICT and spatial structure,
Bekkerman and Gilpin [@bekkerman_high-speed_2013] focused on the role of locally
based information resources using a dataset about the US libraries during the
period $2000-2008$. Their results suggest that internet access increases the
demand and the value of locally accessible information, and such complementarities
are higher in larger metropolitan areas. Anenberg and Kung
[@anenberg_information_2015] also identified a complementary relation between
the internet and consumption variety in cities by focusing on food truck industry
in the US.

\color{black}
Thus, clearly, previous research suggests that the declaration of the “death of distance” has  proven to be premature [@rietveld_transport_2004].
However, the exact impact of digital communications on spatial structure is still an open question.
As Leamer and Storper [@edward_e_leamer_economic_2001] indicate, the internet can affect both centripetal and centrifugal forces.
The only cross country study [@ioannides2008effect] supports a clear substitution effect, whereas most of the above studies have focused either on the US or on some specific cities.
Moreover, most of the above studies examined the complementarity/substitutability question over time periods when the internet and other digital communication technologies were still emerging.
For instance, internet penetration in the US in $2000$, which was the focus for quite a few of the above studies including Ioannides et al. [@ioannides2008effect], was just above $50$ percent, while in $2016$ it reached almost $90$ percent [@pew_internet_pew_2016].
At a global scale internet penetration increased from $7$ to $46$ percent during the same period [@internet_live_stats_internet_2017].
In addition, although email and instant messaging technologies were widespread in the developed world in early $2000s$, network externalities due to mobile internet and online social media were nowhere close to what we are familiar with today.
For instance, Facebook users increased from $1$ million in $2004$ to more than $1.5$ billion in $2015$^[[http://newsroom.fb.com/company-info/](http://newsroom.fb.com/company-info/)].
Hence, it might have been premature for the spatial economic effects of ICT adoption to have been materialized by the time that most of the above studies were conducted.

@@

<!-- \color{blue} -->
The present paper does not address the issue that adoption of ICT may have
different effects on urban dispersion on a national scale, that is how far are
major urban areas, that is, cities, from one another, versus dispersion within major
urban areas, that is urban sprawl. Such an inquiry would require detailed data
about patterns of how urban sprawl may hamper or promote economic interaction,
given advancements in information technology.
We think that data from the Covid-19 pandemic would likely throw light at this
issue in the future, as it is clear that not all economic activities might
continue in an unencumbered manner; see Dingel and Neiman [@dingel2020many].
\color{black}

@@

The multi-country identification strategy is a two-step approach [@ioannides2008effect].
The first step of our methodology is to estimate the
<!-- \color{blue} -->
Pareto exponent
\color{black}
for a broad sample of countries over time.
<!-- \color{blue} -->
The Pareto exponent
\color{black}
is one of the most widely used measures of spatial dispersion with numerous applications in urban economics and economic geography ([see for example @black_urban_2003; @frenken_theoretical_2007; @giesen_zipfs_2011; @rauch_cities_2013; @ioannides_zipfs_2003; @ioannides_walled_2017; @nitsch_zipf_2005).
This an appropriate measure of dispersion because of the extreme heterogeneity of the city size distribution and the very good fits normally obtained with such estimations^[
<!-- \color{blue} -->
We also replicate our analysis using other measures of dispersion in Appendix S1. Our results remain consistent.].
\color{black}
Skipping details pertaining of the suitability of the approach, which may be found in Ioannides et al. [@ioannides2008effect] and Tranos and Ioannides [@tranos2020ict], we proceed with the estimation of the logarithmic form of:

@@

<!-- \color{blue} -->
This leads to an estimate of $\zeta$, known as the Zipf coefficient,
but more appropriately the Pareto exponent, a terminology we will adhere to for
the remained of the text, even  when $\zeta$ is estimated to be near $1$.
The term Zipf coefficient should be reserved for the special case of
$\zeta=1$.
While much of the literature focuses on the Pareto exponent being
close to $1$, several estimations of city size distributions lead to estimates
of $\zeta$ that are not necessarily equal to $1$, or even near it, in which
case we refer to $\zeta$ as the Pareto, or power law, exponent.
Given that our aim here is to estimate $\zeta$ for a number of countries over
time as a measure of dispersion, equation \ref{eq1} describes the rank of city
$i$ in country $c$ in year $t$:
\color{black}

@@

The estimation of equation \ref{eq2} has typically been performed by Ordinary Least Squares (OLS).
Gabaix and Ioannides [@gabaix_evolution_2004] discuss the downward bias of estimates of equation \ref{eq2} using OLS on small samples.
Gabaix and Ibragimov [@gabaix_rank_2011] propose a practical remedy to correct this bias, which we do adopt in this paper:
instead of using the $log$ of $rank$ of a city $i$ in a country $c$ in year $t$, they propose to use the $log(rank-0.5)$, which has indeed been widely adopted.
Researchers working in this area must contend with definitional differences as well as differences in availability of different kinds of data sources.
Definitions of cities differ across countries, for political, administrative and legal reasons^[See [@ioannides2013neighborhoods], Ch. 8, for issues and pitfalls associated with different definitions.].
In contrast to the data drawn from Thomas Brinkhoff’s City Population project [@thomas_brinkhoff_city_2014]^[[http://www.citypopulation.de](http://www.citypopulation.de)], which were  used by Soo [@soo_zipfs_2005], Ioannides et al. [@ioannides2008effect] and Tranos and Ioannides [@tranos2020ict], the multi-country analysis in the present paper is based on the annual population data for urban agglomerations with $300,000$ inhabitants or more from the Department of Economic and Social Affairs of the United Nations [@united_nations_world_2014].
Despite some criticism about the consistency of the urban agglomeration definitions across different countries [@cohen_urban_2004], this is the only available source for yearly, multi-country population data for urban agglomerations [@montgomery_urban_2008; @chen_absolute_2007; @decker_global_2007].
The results based on these data are reported and discussed in [Section 4](#sec4).

@@

<!-- \color{blue} -->

@@

\color{black}
\noindent The model enable us to estimate the effects of a vector $X$ of explanatory variables which are included to account for the spatial structure of country $c$ in year $t$.
The main variables of interest here are internet and digital communications variables that include: internet users per $100$ inhabitants, broadband users per $100$ inhabitants, mobile phone users per $100$ inhabitants, and fixed phone users per $100$ inhabitants.
To address a potential omitted variable bias equation \ref{eq3} includes
<!-- \color{blue} -->
country ($\theta_c$) and year ($\delta_t$) fixed effects;
\color{black}
$e_{ct}$ is the error term.
In addition, vector $X$ includes several control variables, whose descriptive statistics together with those for the other variables used to estimate \ref{eq3} are reported in Table \ref{desc.global}.

@@

\color{black}

@@

In order to capture the micro-dynamics of the urban systems in the two case studies, we follow Batty [@batty_rank_2006] and Havlin [@havlin_distance_1995] and focus on the difference in ranks for individual cities during the study period ^[Changes in ranking of individuals has also been used elsewhere in economics because of its robustness properties. C.f. Chetty and Hendren [@chetty2018impacts].].
Contrary to their approach, we are interested in the real difference in rankings rather than in absolute differences, in order to capture whether or a city improves its ranking within the urban system during the study period.
We then test whether our internet-related variable has an effect on the change in ranking.
Hence, we define the left-hand side (LHS) variable of equation \ref{eq4} as follows:

@@

This section presents the estimation results of \ref{eq3}.
The LHS variable is $\zeta$, the
<!-- \color{blue} -->
Pareto exponent,
\color{black}
as estimated according to the Gabaix and Ibragimov [@gabaix_rank_2011] correction.
The main explanatory variables of interest are provided at the top of Table \ref{ols.global}; namely internet, broadband, mobile and fixed telephony per $100$ habitants, expressed in natural logarithms.
The main variables of interest are introduced successively on their own in the regressions reported in Table \ref{ols.global}.
All regressions include country fixed effects to control for unobserved heterogeneity and a time trend.
In addition, the observations are weighted with the inverse squared standard error of the estimated
<!-- \color{blue} -->
Pareto exponent
\color{black}
(see Table \ref{desc.global}) to address potential noise that is carried over from the first part of our identification strategy.
In regards to the interpretation of the estimated coefficients, given that the
<!-- \color{blue} -->
Pareto exponent
\color{black}
has entered the regression not as an absolute value, but instead as a real number a positive coefficient for a RHS variable indicates an impact towards the decrease of the spatial dispersion of population. In other words, a positive coefficient indicates an effect towards less uniform city sizes that is more dispersion of city sizes.
The latter is indicative of enhancement of agglomeration economies because of the expansion of digital technologies.

@@

Table \ref{ols.global} and \ref{2sls.global} report the estimation results based on the UN urban agglomeration data.
Starting with the former, we note that an agglomerative effect is only detected for the internet users and while a marginally opposing result can be seen for broadband users as indicated by the significant coefficients in columns (1) and (2).
For the telephony variables the estimation of \ref{eq3} did not yield statistically significant coefficients.
Before discussing further these results and the effects of the other control variables, we need to highlight that the main challenge of estimating equation \ref{eq3} is the potential endogenous nature of the share of internet users which might prevent us from being able to infer a truly causal effect.
Endogeneity might be an issue here as spatial structure, which is proxied by the
<!-- \color{blue} -->
Pareto exponent,
\color{black}
might be affected by another source, which also affects internet penetration.
For instance, economic development might affect the concentration of population in large cities and at the same time enable more people to go online.
If we do not address this issue, the coefficient for the main variable of interest will capture potential effects that internet penetration has on spatial structure, but also potential reverse causality effects that spatial structure might generate on internet penetration.
To overcome this potential problem, Table \ref{2sls.global} reports estimates of equation \ref{eq3} using two-stage least squares (2SLS) with instrumental variables (IVs).
The latter are variables which are correlated with our endogenous variables, but do not influence current spatial structure.
Such an approach will enable us to estimate the causal effect -- if any -- of the internet and digital communications on spatial structure.
At a first stage, our endogenous variables are regressed against the IVs.
Then, the predicted values of the endogenous variables based on the IVs and the other control variables are used instead of the endogenous variable to estimate eq. \ref{eq3}.
A significant effect will verify the causal impact of the internet and digital communication usage on spatial structure.

@@

The main challenge for such an exercise is to find a valid IV.
In our case, the challenge is even bigger as we need to find a variable which is not only fit for purpose, but it also contain as few missing values as possible in order to retain the same number of observations as the OLS models.
This is particularly difficult for a diverse sample of countries as the one we are dealing with here.
We propose here the percentage of female participation in the labor force, which is directly related to digital infrastructure, but not to spatial structure (see Table \ref{desc.global} for the descriptive statistics).
<!-- \color{blue} -->
The literature provides empirical and theoretical evidence that ICT are correlated with female participation in labour markets mostly because they enable teleworking [@dettling2017; @arntz2019working; @watson2018digitalisation; @Lechman2020; @Valberg2020]
\color{black}
At the same time, female labor participation is not related with spatial structure especially given the diverse sample of countries.
<!-- Firstly, the correlation coefficient between the IV and the Zipf coefficient is very low ($-0.14$). -->
Although one might think that increased female labor participation may lead to relocation of households to large cities and therefore affect the spatial structure of cities, such a process varies a lot within our dataset.
Even if this might be true for developed countries with mature urban systems for which a location within a large urban center might provide opportunities for both male and female workers to find jobs, our dataset also includes countries, $41$ per cent of the GDP of which is attributed to primary activities (see Table \ref{desc.global}).
So, female labor participation might also be related to economic activities located outside large urban centers.
In terms of the relevant tests presented in Table \ref{2sls.global} the weak identification first stage F-test exceeds by far all the Stock-Yogo critical values for all, but the mobile telephony regression.
Hence, we are confident to interpret the effects of internet, broadband internet and fixed telephony usage.

The estimations presented in Table \ref{2sls.global} indicate that increases in internet and broadband internet usage have resulted to a decrease of the spatial dispersion of population for the time period and for the panel of countries included in our data.
In other words, increases in internet and digital communications resulted to national urban systems which are less uniform in terms of city sizes and are characterized by higher population concentrations in larger cities.
The exact opposite appears to happen for fixed telephony, which our results indicate that has led to an increase of spatial dispersion of population.
<!-- \color{blue} -->
We expect that the IV estimates capture the local average treatment effect
[@angrist1995identification] of these labour markets, which are characterised by
more active engagement with the internet, and, therefore, experience a larger effect
on the Pareto exponents.
\color{black}
Importantly, our identification strategy enables us to address potential reverse causality issues and treat the results of Table \ref{2sls.global} as causal. Hence, the main finding of our multi-country, global analysis based on data for urban agglomerations of at least 300,000 inhabitants is that internet and broadband internet usage appear to act in favor of agglomeration economies and result to urban systems with more dominant cities on the top of the urban hierarchy.
On the contrary, fixed telephony usage acted in favour of centrifugal forces.
Our identification strategy does not enable us to make similar interpretations for mobile telephony usage.

@@

In total, the results indicate that our measures of internet and telephony penetration have further enhanced agglomeration forces, at least for large urban agglomerations, while the opposite happens for fixed telephony.
These results are robust against endogeneity issues, but are limited to urban agglomeration included in our data.
Indeed, for the estimation of the
<!-- \color{blue} -->
Pareto exponent
\color{black}
we only included urban agglomeration of $300,000$ inhabitants or more due to data availability.
Hence, the above estimations cannot verify whether such an effect is also valid for smaller cities.
Interestingly, the above results regarding the role of fixed telephony are in agreement with previous results from [@ioannides2008effect] [@tranos2020ict] which used a smaller panel of countries but without constraints regarding the city size.
However, our results do not support the findings of [@tranos2020ict] regarding internet usage.

<!-- \color{blue} -->
It is indeed puzzling that the results on Table \ref{2sls.global} that the
effects on urban concentration for internet users and broadband users, on the
one hand, and mobile and fixed telephony users are in opposite directions.
We wish to broach this subject in the context of some related literature,
namely [@calabrese2011interplay] and [@buchel2020cities].
The latter supports the notion that ICT favors agglomeration in the sense that
in spite of their convenience modern technologies do no make up for the need of
proximity in  social interactions.
The former supports the notion people who communicate more frequently are likely
to be near one another.
ICT, as viewed through such granular studies of individuals' communications
do not make up for physical distance; individuals nearer one another
communicate more.  
\color{black}

>>>>

Urban economics suggests that a key source of agglomeration externalities is proximity, which  facilitates interaction and knowledge spillovers [@rosenthal_evidence_2004].
Hence, face-to-face interactions and the implied knowledge spillovers within cities flourish because of lower transportation and interaction costs [@storper_buzz_2004].
However, ICT have the capacity to directly affect this process by further reducing transportation cost [@mccann_globalization_2008].
In essence, the internet and digital communications promote further learning and matching, which are at the heart of the micro-foundations of agglomeration economies identified by Duranton and Puga [@duranton_micro-foundations_2004].
Web-based applications such as Massive Open Online Courses decrease the need for co-location of actors in order to participate in formal learning activities.
Furthermore, online social media such as LinkedIn can enhance the probability of matching and indeed the quality of matches even within cities.
This is why LinkedIn has been identified as "the largest professional matchmaker site in the world" [[@van_dijck_you_2013] p. 207; [@gee2019more]].
Therefore, it is natural to ask whether the widespread adoption of ICT could offset the benefits of  immediate physical proximity and result in more dispersed spatial structures, or further reinforce agglomeration  externalities and lead to more concentrated spatial structures.

@@

Kolko [@kolko_death_2000] uses internet diffusion data and identifies a clear complementary link between internet usage and city size.
Interestingly, he identifies higher internet domain densities in remote cities which indicates a substitution effect of the internet for longer-distance non-electronic communications. His results are consistent for different measures of internet diffusion (internet domain density and internet take-up).
Sinai and Waldfogel [@todd_sinai_geography_2004] approach the same question from the consumers’ point of view and find support for complementarity between internet and cities.
They study the link between market size and locally targeted online content and find more online local content in  larger markets.
However, they also obtain evidence for a substitution effect: holding local online content constant, market size has a negative effect on individual connectivity.
Forman et al. [@chris_forman_how_2005] examines whether commercial internet adoption is higher in cities than in rural areas. While the former would indicate a complementarity between internet adoption and cities, the latter would reflect a substitutability relation, according to which the internet is used as a means to offset costs and lack of opportunities related to peripheriality.
Their results indicate that despite internet adoption by firms with more than 100 employees being greater in smaller urban agglomerations, the adoption of more sophisticated internet-based applications is positively related with city size in $2000$.
Sohn, Kim, and Hewings [@jungyul_sohn_information_2003] compare how information technologies are related to urban spatial structure for Chicago and Seoul.
Although they find a clear complementary link for Chicago, this is not the case for Seoul, where information technologies contribute to a more dispersed spatial pattern.
Focusing on the municipalities in the province of Barcelona, Pons-Novell and Viladecans-Marsal [@pons-novell_cities_2006] find a complementary link between individual internet take-up and off-line commercial offerings, but their results cannot safely reject a substitution effect.
Craig, Hoang, and Kohlhase [@craig_does_2016] focus on internet take-up rates across US states during the period $2000-2011$.
Their analysis provides suggestive evidence of a complementary role that internet connectivity performs on urban living.
Ioannides et al. [@ioannides2008effect] use country-level data on city size distributions to examine the impact of fixed line telephony on urban structure $1980-2000$.
Using a panel dataset of spatial dispersion measures, they find robust evidence that an increase in the number of telephone lines per capita encourages the spatial dispersion of population in that they lead to a more concentrated distribution of city sizes.
In addition, by the end of their study period the internet has come into use, but their results with internet usage is more speculative but do show that it goes in the same direction.

Focusing on rural areas, Partridge et al. [@mark_d_partridge_employment_2008] find no evidence that rural distance penalties in the US have substantially changed since $1970s$ indicating that technological changes including the internet and digital communications have not affected spatial structure. Also interesting are the findings of a more recent study by Kim and Orazem [@kim_broadband_2016] on the economic effects of broadband internet in rural areas. They identify a positive effect on new firm location decisions, with the effect being higher in rural areas with larger population and in rural areas adjacent to a metropolitan area. These results suggest a complementary relation between the internet and agglomeration economies.

Moving beyond modelling the direct relationship between ICT and spatial structure,
Bekkerman and Gilpin [@bekkerman_high-speed_2013] focused on the role of locally
based information resources using a dataset about the US libraries during the
period $2000-2008$. Their results suggest that internet access increases the
demand and the value of locally accessible information, and such complementarities
are higher in larger metropolitan areas. Anenberg and Kung
[@anenberg_information_2015] also identified a complementary relation between
the internet and consumption variety in cities by focusing on food truck industry
in the US.

Thus, clearly, previous research suggests that the declaration of the “death of distance” has  proven to be premature [@rietveld_transport_2004].
However, the exact impact of digital communications on spatial structure is still an open question.
As Leamer and Storper [@edward_e_leamer_economic_2001] indicate, the internet can affect both centripetal and centrifugal forces.
The only cross country study [@ioannides2008effect] supports a clear substitution effect, whereas most of the above studies have focused either on the US or on some specific cities.
Moreover, most of the above studies examined the complementarity/substitutability question over time periods when the internet and other digital communication technologies were still emerging.
For instance, internet penetration in the US in $2000$, which was the focus for quite a few of the above studies including Ioannides et al. [@ioannides2008effect], was just above $50$ percent, while in $2016$ it reached almost $90$ percent [@pew_internet_pew_2016].
At a global scale internet penetration increased from $7$ to $46$ percent during the same period [@internet_live_stats_internet_2017].
In addition, although email and instant messaging technologies were widespread in the developed world in early $2000s$, network externalities due to mobile internet and online social media were nowhere close to what we are familiar with today.
For instance, Facebook users increased from $1$ million in $2004$ to more than $1.5$ billion in $2015$ [@fb].
Hence, it might have been premature for the spatial economic effects of ICT adoption to have been materialized by the time that most of the above studies were conducted.

@@

The present paper does not address the issue that adoption of ICT may have
different effects on urban dispersion on a national scale, that is how far are
major urban areas, that is, cities, from one another, versus dispersion within major
urban areas, that is urban sprawl. Such an inquiry would require detailed data
about patterns of how urban sprawl may hamper or promote economic interaction,
given advancements in information technology.
We think that data from the Covid-19 pandemic would likely throw light at this
issue in the future, as it is clear that not all economic activities might
continue in an unencumbered manner; see Dingel and Neiman [@dingel2020many].

@@

The multi-country identification strategy is a two-step approach [@ioannides2008effect].
The first step of our methodology is to estimate the
Pareto exponent
for a broad sample of countries over time.
The Pareto exponent is one of the most widely used measures of spatial dispersion with numerous applications in urban economics and economic geography ([see for example @black_urban_2003; @frenken_theoretical_2007; @giesen_zipfs_2011; @rauch_cities_2013; @ioannides_zipfs_2003; @ioannides_walled_2017; @nitsch_zipf_2005).
This an appropriate measure of dispersion because of the extreme heterogeneity of the city size distribution and the very good fits normally obtained with such estimations.
We also replicate our analysis using other measures of dispersion in Appendix S1. Our results remain consistent.
Skipping details pertaining of the suitability of the approach, which may be found in Ioannides et al. [@ioannides2008effect] and Tranos and Ioannides [@tranos2020ict], we proceed with the estimation of the logarithmic form of:

@@

This leads to an estimate of $\zeta$, known as the Zipf coefficient,
but more appropriately the Pareto exponent, a terminology we will adhere to for
the remained of the text, even  when $\zeta$ is estimated to be near $1$.
The term Zipf coefficient should be reserved for the special case of
$\zeta=1$.
While much of the literature focuses on the Pareto exponent being
close to $1$, several estimations of city size distributions lead to estimates
of $\zeta$ that are not necessarily equal to $1$, or even near it, in which
case we refer to $\zeta$ as the Pareto, or power law, exponent.
Given that our aim here is to estimate $\zeta$ for a number of countries over
time as a measure of dispersion, equation \ref{eq1} describes the rank of city
$i$ in country $c$ in year $t$:

@@

The estimation of equation \ref{eq2} has typically been performed by Ordinary Least Squares (OLS).
Gabaix and Ioannides [@gabaix_evolution_2004] discuss the downward bias of estimates of equation \ref{eq2} using OLS on small samples.
Gabaix and Ibragimov [@gabaix_rank_2011] propose a practical remedy to correct this bias, which we do adopt in this paper:
instead of using the $log$ of $rank$ of a city $i$ in a country $c$ in year $t$, they propose to use the $log(rank-0.5)$, which has indeed been widely adopted.
Researchers working in this area must contend with definitional differences as well as differences in availability of different kinds of data sources.
Definitions of cities differ across countries, for political, administrative and legal reasons [see [@ioannides2013neighborhoods], Ch. 8, for issues and pitfalls associated with different definitions].
In contrast to the data drawn from Thomas Brinkhoff’s City Population project [@citypop], 
which were  used by Soo [@soo_zipfs_2005], Ioannides et al. [@ioannides2008effect] and Tranos and Ioannides [@tranos2020ict], the multi-country analysis in the present paper is based on the annual population data for urban agglomerations with $300,000$ inhabitants or more from the Department of Economic and Social Affairs of the United Nations [@united_nations_world_2014].
Despite some criticism about the consistency of the urban agglomeration definitions across different countries [@cohen_urban_2004], this is the only available source for yearly, multi-country population data for urban agglomerations [@montgomery_urban_2008; @chen_absolute_2007; @decker_global_2007].
The results based on these data are reported and discussed in [Section 4](#sec4).

@@

@@

\noindent The model enable us to estimate the effects of a vector $X$ of explanatory variables which are included to account for the spatial structure of country $c$ in year $t$.
The main variables of interest here are internet and digital communications variables that include: internet users per $100$ inhabitants, broadband users per $100$ inhabitants, mobile phone users per $100$ inhabitants, and fixed phone users per $100$ inhabitants.
To address a potential omitted variable bias equation \ref{eq3} includes
country ($\theta_c$) and year ($\delta_t$) fixed effects;
$e_{ct}$ is the error term.
In addition, vector $X$ includes several control variables, whose descriptive statistics together with those for the other variables used to estimate \ref{eq3} are reported in Table \ref{desc.global}.

@@

@@

In order to capture the micro-dynamics of the urban systems in the two case studies, we follow Batty [@batty_rank_2006] and Havlin [@havlin_distance_1995] and focus on the difference in ranks for individual cities during the study period.
Changes in ranking of individuals have also been used elsewhere in economics because 
of its robustness properties. C.f. Chetty and Hendren [@chetty2018impacts].
Contrary to their approach, we are interested in the real difference in rankings rather than in absolute differences, in order to capture whether or a city improves its ranking within the urban system during the study period.
We then test whether our internet-related variable has an effect on the change in ranking.
Hence, we define the left-hand side (LHS) variable of equation \ref{eq4} as follows:

@@

This section presents the estimation results of \ref{eq3}.
The LHS variable is $\zeta$, the Pareto exponent,
as estimated according to the Gabaix and Ibragimov [@gabaix_rank_2011] correction.
The main explanatory variables of interest are provided at the top of Table \ref{ols.global}; namely internet, broadband, mobile and fixed telephony per $100$ habitants, expressed in natural logarithms.
The main variables of interest are introduced successively on their own in the regressions reported in Table \ref{ols.global}.
All regressions include country fixed effects to control for unobserved heterogeneity and a time trend.
In addition, the observations are weighted with the inverse squared standard error of the estimated
Pareto exponent (see Table \ref{desc.global}) to address potential noise that is carried over from the first part of our identification strategy.
In regards to the interpretation of the estimated coefficients, given that the Pareto exponent
has entered the regression not as an absolute value, but instead as a real number a positive coefficient for a RHS variable indicates an impact towards the decrease of the spatial dispersion of population. In other words, a positive coefficient indicates an effect towards less uniform city sizes that is more dispersion of city sizes.
The latter is indicative of enhancement of agglomeration economies because of the expansion of digital technologies.

@@

Table \ref{ols.global} and \ref{2sls.global} report the estimation results based on the UN urban agglomeration data.
Starting with the former, we note that an agglomerative effect is only detected for the internet users and while a marginally opposing result can be seen for broadband users as indicated by the significant coefficients in columns (1) and (2).
For the telephony variables the estimation of \ref{eq3} did not yield statistically significant coefficients.
Before discussing further these results and the effects of the other control variables, we need to highlight that the main challenge of estimating equation \ref{eq3} is the potential endogenous nature of the share of internet users which might prevent us from being able to infer a truly causal effect.
Endogeneity might be an issue here as spatial structure, which is proxied by the
Pareto exponent,
might be affected by another source, which also affects internet penetration.
For instance, economic development might affect the concentration of population in large cities and at the same time enable more people to go online.
If we do not address this issue, the coefficient for the main variable of interest will capture potential effects that internet penetration has on spatial structure, but also potential reverse causality effects that spatial structure might generate on internet penetration.
To overcome this potential problem, Table \ref{2sls.global} reports estimates of equation \ref{eq3} using two-stage least squares (2SLS) with instrumental variables (IVs).
The latter are variables which are correlated with our endogenous variables, but do not influence current spatial structure.
Such an approach will enable us to estimate the causal effect -- if any -- of the internet and digital communications on spatial structure.
At a first stage, our endogenous variables are regressed against the IVs.
Then, the predicted values of the endogenous variables based on the IVs and the other control variables are used instead of the endogenous variable to estimate eq. \ref{eq3}.
A significant effect will verify the causal impact of the internet and digital communication usage on spatial structure.

@@

The main challenge for such an exercise is to find a valid IV.
In our case, the challenge is even bigger as we need to find a variable which is not only fit for purpose, but it also contain as few missing values as possible in order to retain the same number of observations as the OLS models.
This is particularly difficult for a diverse sample of countries as the one we are dealing with here.
We propose here the percentage of female participation in the labor force, which is directly related to digital infrastructure, but not to spatial structure (see Table \ref{desc.global} for the descriptive statistics).
The literature provides empirical and theoretical evidence that ICT are correlated with female participation in labour markets mostly because they enable teleworking [@dettling2017; @arntz2019working; @watson2018digitalisation; @Lechman2020; @Valberg2020]
At the same time, female labor participation is not related with spatial structure especially given the diverse sample of countries.
<!-- Firstly, the correlation coefficient between the IV and the Zipf coefficient is very low ($-0.14$). -->
Although one might think that increased female labor participation may lead to relocation of households to large cities and therefore affect the spatial structure of cities, such a process varies a lot within our dataset.
Even if this might be true for developed countries with mature urban systems for which a location within a large urban center might provide opportunities for both male and female workers to find jobs, our dataset also includes countries, $41$ per cent of the GDP of which is attributed to primary activities (see Table \ref{desc.global}).
So, female labor participation might also be related to economic activities located outside large urban centers.
In terms of the relevant tests presented in Table \ref{2sls.global} the weak identification first stage F-test exceeds by far all the Stock-Yogo critical values for all, but the mobile telephony regression.
Hence, we are confident to interpret the effects of internet, broadband internet and fixed telephony usage.

The estimations presented in Table \ref{2sls.global} indicate that increases in internet and broadband internet usage have resulted to a decrease of the spatial dispersion of population for the time period and for the panel of countries included in our data.
In other words, increases in internet and digital communications resulted to national urban systems which are less uniform in terms of city sizes and are characterized by higher population concentrations in larger cities.
The exact opposite appears to happen for fixed telephony, which our results indicate that has led to an increase of spatial dispersion of population.
We expect that the IV estimates capture the local average treatment effect
[@angrist1995identification] of these labour markets, which are characterised by
more active engagement with the internet, and, therefore, experience a larger effect
on the Pareto exponents.
Importantly, our identification strategy enables us to address potential reverse causality issues and treat the results of Table \ref{2sls.global} as causal. Hence, the main finding of our multi-country, global analysis based on data for urban agglomerations of at least 300,000 inhabitants is that internet and broadband internet usage appear to act in favor of agglomeration economies and result to urban systems with more dominant cities on the top of the urban hierarchy.
On the contrary, fixed telephony usage acted in favour of centrifugal forces.
Our identification strategy does not enable us to make similar interpretations for mobile telephony usage.

@@

In total, the results indicate that our measures of internet and telephony penetration have further enhanced agglomeration forces, at least for large urban agglomerations, while the opposite happens for fixed telephony.
These results are robust against endogeneity issues, but are limited to urban agglomeration included in our data.
Indeed, for the estimation of the
Pareto exponent
we only included urban agglomeration of $300,000$ inhabitants or more due to data availability.
Hence, the above estimations cannot verify whether such an effect is also valid for smaller cities.
Interestingly, the above results regarding the role of fixed telephony are in agreement with previous results from [@ioannides2008effect] [@tranos2020ict] which used a smaller panel of countries but without constraints regarding the city size.
However, our results do not support the findings of [@tranos2020ict] regarding internet usage.

It is indeed puzzling that the results on Table \ref{2sls.global} that the
effects on urban concentration for internet users and broadband users, on the
one hand, and mobile and fixed telephony users are in opposite directions.
We wish to broach this subject in the context of some related literature,
namely [@calabrese2011interplay] and [@buchel2020cities].
The latter supports the notion that ICT favors agglomeration in the sense that
in spite of their convenience modern technologies do no make up for the need of
proximity in  social interactions.
The former supports the notion people who communicate more frequently are likely
to be near one another.
ICT, as viewed through such granular studies of individuals' communications
do not make up for physical distance; individuals nearer one another
communicate more.
